---
title: "Heart Disease Analysis"
author: "German Batuista (gbauti5@illinois.edu)"
date: "04/24/2021"
output:
  html_document: 
    theme: default
    toc: yes
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center')
```

```{r, load-packages, include = FALSE}
# load packages
```

```{r read-full-data, warning = FALSE, message = FALSE}
# read full data
hd = readr::read_csv("data/hd.csv")
```

***

## Abstract

> Abstract text goes here.

DO THIS LAST AFTER THE ANALYSIS.
SHORT AND SWEET.
***

## Introduction

Write your introduction here.

***

## Methods

Describe the methods used in this section.

### Data

Specifically describe the data and how it is used here.

### Modeling

Discuss your modeling and validation strategy here.



FOR SIMPLICITY, WE WILL USE THE CLEAN DATA. Check `check-data` for informaiton on how hd_clean was obtained.
```{r}
trn_idx = sample(nrow(hd_clean), size = nrow(hd_clean) * 0.80, replace = FALSE)
hd_trn = hd_clean[trn_idx, ]
hd_tst = hd_clean[-trn_idx, ]

summary(hd_trn)
```


***

## Results

State your results here.

***

## Discussion

Discuss your results here.

***

## Appendix

Place potential additional information here.
